Economic Theory of Fertility Explains Modern Natural Selection

### 2020-11-16 why-negative-selection.Rmd
tweaks from abdel comments

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are larger among low-income and less educated people, younger parents, and people not living with a partner. Overall effect sizes are substantially bigger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

pretty_names <- function (names) {
  pretty <- c(
    ADHD_2017               = "ADHD",
    age_at_menarche         = "Age at menarche",
    age_at_menopauze        = "Age at menopause",
    agreeableness           = "Agreeableness",
    ai_substance_use        = "Age at smoking initiation",
    alcohol_schumann        = "Alcohol use",
    alzheimer               = "Alzheimer",
    autism_2017             = "Autism",
    bipolar                 = "Bipolar",
    bmi_combined            = "BMI",
    body_fat                = "Body Fat",
    caffeine                = "Caffeine",
    cannabis                = "Cannabis (ever vs. never)",
    cognitive_ability       = "Cognitive Ability",
    conscientiousness       = "Conscientiousness",
    coronary_artery_disease = "Coronary Artery Disease",
    cpd_substance_use       = "Cigarettes per day",
    diagram_T2D             = "Type 2 Diabetes",
    dpw_substance_use       = "Drinks per week",
    EA2_noUKB               = "Educ. attainment 2 (no UK)",
    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
    eating_disorder         = "Eating disorder",
    extraversion            = "Extraversion",
    height_combined         = "Height",
    hip_combined            = "Hip circumference",
    MDD_PGC2_noUKB          = "Major Depressive Disorder",
    neuroticism             = "Neuroticism",
    openness                = "Openness",
    sc_substance_use        = "Smoking cessation",
    SCZ2                    = "Schizophrenia",
    si_substance_use        = "Smoking initiation",
    wc_combined             = "Waist circumference",
    whr_combined            = "Waist-hip ratio"
  )

@@

Geneticists have long hypothesized that natural selection is taking
place in modern human populations. Recent work confirms this using
genome-wide analysis [@Barban_2016; @beauchamp2016genetic;
@kong2017selection; @sanjak2018evidence]. In particular, genetic
variants associated with higher educational attainment are being
selected against, although effect sizes appear small.

As yet we know little about the *mechanisms* behind natural selection in
contemporary populations. This paper uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We
extend the analysis over two generations, using data on respondents'
number of siblings as well as their number of children to reconstruct
two generations of family size. This is interesting because consistent
natural selection over multiple generations could lead to substantive
effects in the long run. Most importantly, we examine the strength of
natural selection in different subgroups of the population, in order to
uncover patterns that can help illuminate the mechanisms involved.

@@

We next focus on variables related to mate search and household
formation. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Results are unchanged if we control for
respondents' age (appendix Figure
\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
they were living with a spouse or partner at the time of interview,
effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within categories, and
shifting respondents between categories.

@@

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, and that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
no role in the decision to leave education, since people are not yet earning
wages.[^no-borrow] Whilst in the labour market, higher earners will have more
children (i.e. income effects dominate). These assumptions will lead to the
pattern we observe: earnings-increasing variants lead to later first births (and
fewer children overall), but more children for any given AFLB.

@@

```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

@@

```{r plot-res-wt, fig.cap = "Effects of polygenic scores on fertility. Weighted regressions."}

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents"}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in effect sizes between parents of early and late born respondents (regressions on number of siblings)")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born respondents (regressions on number of children)")
```

@@

```{r plot-regs-controlled-pcs, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}

@@

```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children. Absolute effect sizes are plotted. Each dot represents one bivariate regression. Points are jittered on the Y axis.", fig.height = 2}

@@

```{r plot-n-partners-controlled, fig.cap = "Effect sizes on number of siblings interacted with number of sexual partners (median split at 3), respondent's age and age squared."}
drake::loadd(res_partners_controlled)
res_partners_controlled %<>%       
      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects controlling for parents' age at birth, eldest siblings"}

@@

```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's age", 
          "moth_age_birth" = "Mother's age"
        )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs()

@@

```{r plot-rgs-by-effect-size, fig.cap = "Effect size on n siblings by genetic correlation with EA3"}

@@

\FloatBarrier
## Number of children

Figure \@ref(fig:plot-n-children-bar) shows the full distribution of
number of children born for different ventiles of the EA3 polygenic
score. The strongest relationship seems to be for having 0 children
versus 1 or more.

```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
famhist %>% 
      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
      mutate(
        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
        "N Children"  = santoku::chop(n_children, 0:4, labels = c(0:3, "4+")),
        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
        ) %>% 
      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
        facet_wrap(vars(Sex)) +
        scale_fill_viridis_d(option = "A") +
        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
        theme(legend.position = "bottom")

@@

```{r plot-together, fig.cap = "Partial correlations with number of children"}

>>>>

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are larger among low-income and less educated people, younger parents, and people not living with a partner. Overall effect sizes are substantially bigger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

pretty_names <- function (names) {
  pretty <- c(
    ADHD_2017               = "ADHD",
    age_at_menarche         = "Age at menarche",
    age_at_menopauze        = "Age at menopause",
    agreeableness           = "Agreeableness",
    ai_substance_use        = "Age at smoking initiation",
    alcohol_schumann        = "Alcohol use",
    alzheimer               = "Alzheimer",
    autism_2017             = "Autism",
    bipolar                 = "Bipolar",
    bmi_combined            = "BMI",
    body_fat                = "Body Fat",
    caffeine                = "Caffeine",
    cannabis                = "Cannabis (ever vs. never)",
    cognitive_ability       = "Cognitive Ability",
    conscientiousness       = "Conscientiousness",
    coronary_artery_disease = "Coronary Artery Disease",
    cpd_substance_use       = "Cigarettes per day",
    diagram_T2D             = "Type 2 Diabetes",
    dpw_substance_use       = "Drinks per week",
    EA2_noUKB               = "Educ. attainment 2 (no UKBB)",
    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
    eating_disorder         = "Eating disorder",
    extraversion            = "Extraversion",
    height_combined         = "Height",
    hip_combined            = "Hip circumference",
    MDD_PGC2_noUKB          = "Major Depressive Disorder",
    neuroticism             = "Neuroticism",
    openness                = "Openness",
    sc_substance_use        = "Smoking cessation",
    SCZ2                    = "Schizophrenia",
    si_substance_use        = "Smoking initiation",
    wc_combined             = "Waist circumference",
    whr_combined            = "Waist-hip ratio"
  )

@@

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Geneticists
have long hypothesized that natural selection is taking place in modern human
populations. Recent work confirms this using genome-wide analysis [@Barban_2016;
@beauchamp2016genetic;
@kong2017selection; @sanjak2018evidence]. In particular, genetic variants
associated with higher educational attainment are being selected against,
although effect sizes appear small.

As yet we know little about the *mechanisms* behind natural selection in
contemporary human populations. This study uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We extend
the analysis over two generations, using data on respondents' number of siblings
as well as their number of children. This is interesting because consistent
natural selection over multiple generations could lead to substantive effects in
the long run. Most importantly, we examine reproductive rates in different
subgroups of the population, in order to uncover patterns that can help
illuminate the mechanisms behind modern natural selection.

@@

We next focus on variables related to mate search and household
formation. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Results are unchanged if we control for
respondents' age (appendix Figure
\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
they were living with a spouse or partner at the time of interview,
effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within each category, and changing
the number of children by shifting respondents between categories.

@@

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, i.e. skills and characteristics that are valuable in the labour market.
Suppose also that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
no role in the decision to leave education, since people are not yet earning
wages.[^no-borrow] Whilst in the labour market, higher earners will have more
children (i.e. income effects dominate). These assumptions will lead to the
pattern we observe: earnings-increasing variants will predict later first births (and
fewer children overall), but they will also predict more children within any given
age at first live birth.

@@

```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}

@@

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions."}

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents."}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings)")

@@

```{r tbl-children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in selection effects between early and late born respondents (regressions on number of children)")
```

@@

```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using residualized polygenic scores on number of siblings/children."}

@@

```{r plot-pc-regs, fig.cap = "Selection effects of principal components of genetic data. Absolute effect sizes are plotted. Each dot represents one bivariate regression. Points are jittered on the Y axis.", fig.height = 2}

@@

```{r plot-n-partners-controlled, fig.cap = "Selection effects by number of sexual partners (median split at 3), controlling for respondent's age and age squared."}
drake::loadd(res_partners_controlled)
res_partners_controlled %<>%       
      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth."}

@@

```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth."}

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings."}

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs()

@@

```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}

@@

@@

```{r plot-together, fig.cap = "Partial correlations of PGS with fertility, from a single regression."}
